---
title: An R Markdown document converted from "0_Export_from_UI.ipynb"
output: html_document
---

# Export data from User Interface (UI) to analysis workspace
This is a tutorial notebook that walks through the process of exporting selected data from the *NHLBI BioData Catalyst® (BDC) Powered by PIC-SURE* User Interface, or UI, into an analysis workspace. This is done using the *BDC Powered by PIC-SURE (BDC-PIC-SURE)* Application Programming Interface, or API.

------- 
 
## Introduction to exporting data into an analysis workspace with PIC-SURE

Two things are needed to export data into an analysis workspace:
1. Personalized access token: a user-specific token that tells PIC-SURE which studies a user is authorized to access
2. Query ID: a token that describes the specific query that was built in the UI, for example, if a user has selected females with body mass index between 18 and 30 from the ARIC study

Using these two components, the API can be used to export the selected data into the analysis workspace (in this case, where this Jupyter Notebook is being run). 

## Step 1: Getting your user-specific security token
**Before running this notebook, please be sure to review the "Get your security token" documentation, which exists in the [`README.md` file](../README.md). It explains about how to get a security token, which is mandatory to access the databases.**

## Step 2: Setting up your notebook

### Pre-requisites for the notebook
* R 4.1 or later

### Install packages to connect to the PIC-SURE API
The first step to using the PIC-SURE API is to install the packages needed. The following code installs the PIC-SURE API components from GitHub, specifically:
* PIC-SURE Client
* PIC-SURE Adapter
* *BDC-PIC-SURE* Adapter

**Note that if you are using the dedicated PIC-SURE environment within the *BDC Powered by Seven Bridges (BDC-Seven Bridges)* platform, the necessary packages have already been installed.**

*Troubleshooting: if you encounter an error while installing packages or connecting to a PIC-SURE resource, try restarting your kernel and trying again.*

```{r}
### Uncomment this code if you are not using the PIC-SURE environment in *BDC-Seven Bridges*, or if you do not have all the necessary dependencies installed.
#install.packages("devtools")
```

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
devtools::install_github("hms-dbmi/pic-sure-r-adapter-hpds", ref="query_v3", force=T, quiet=FALSE)
library(dplyr)
```

## Connecting to a PIC-SURE resource
The following is required to get access to data through the PIC-SURE API: 
- User-specific security token

The following code connects to the *BDC-PIC-SURE* platform using the token saved in `token.txt`.

If you have not already retrieved your user-specific token, please refer to the "Get your security token" section of the `README.md` file and the `Workspace_setup.ipynb` file.

```{r}
library(picsure)
token_file <- "token.txt"
my_token <- readLines(token_file, warn = FALSE)[1]
session <- picsure::connect(platform = picsure::Platform$BDC_AUTHORIZED, token = my_token)
```

## Step 3: Export data from a query built in the PIC-SURE UI using the Query ID

You are able to retrieve the results of a query that you have previously built using the [PIC-SURE UI](https://picsure.biodatacatalyst.nhlbi.nih.gov/). After you have built your query and filtered to your cohort of interest, click the "Prepare Data for Analysis" button and copy the Query ID.

Paste the copied Query ID into the `queryID` variable in the cell below, then run `picsure::runQueryByID` to retrieve the results as a data frame. A real Query ID from the UI is required to execute this cell.


```{r}
# To run this using your notebook you must replace it with the ID value of a query that you have run.
queryID <- 'Paste your Query ID here'
```

```{r}
# `queryID` is copied from the PIC-SURE UI ("Copy Query ID").
results <- picsure::runQueryByID(session, queryID, type = picsure::QueryType$PARTICIPANT)

# View the first few records
head(results)
```

The data has now been exported as a dataframe saved as `results` and is ready for analysis.

## Bonus: Inspect a query built in the PIC-SURE UI
You can load a saved query from the PIC-SURE UI into R to inspect it or use it as the starting point for a new query composed in R.

`picsure::loadQueryByID` returns an R object representing the query; use `str()` to examine its structure. To run modified versions of the query, compose new clause and group objects in R and pass them to `picsure::buildQuery` and `picsure::runQuery` (see the editing example further below).

```{r}
# Load the saved query object from the UI (uses the same queryID as above).
query <- picsure::loadQueryByID(session, queryID)

# Inspect the structure of the returned query object.
str(query)
```

### Inspecting a loaded query in v3

In the v3 adapter there is no live mutable query object that exposes `$requiredFields` or `$categoryFilters` slots. Instead, the query structure is reflected through standard R objects:

- `str(query)` on the result of `picsure::loadQueryByID` shows the full parsed query structure.
- To build or modify a query in R, construct clause objects with `picsure::buildClause`, optionally group them with `picsure::buildClauseGroup`, and assemble them into a query with `picsure::buildQuery`. You can then inspect the resulting R list/object with `str()` before running it.

See the editing example in the next section and the `1_PICSURE_API_101` notebook for detailed examples.