---
title: An R Markdown document converted from "1_PICSURE_API_101.ipynb"
output: html_document
---

# Introduction to the *NHLBI BioData Catalyst® (BDC) Powered by PIC-SURE* API

This is a tutorial notebook aimed to get the user quickly up and running with the R PIC-SURE API. 

## PIC-SURE R API
### What is PIC-SURE?

As part of the *NHLBI BioData Catalyst® (BDC)* ecosystem, the Patient Information Commons Standard Unification of Research Elements (PIC-SURE) platform has been integrating clinical and genomic datasets from multiple TOPMed and TOPMed related studies funded by the National Heart Lung and Blood Institute (NHLBI).

Original data exposed through PIC-SURE API encompasses a large heterogeneity of data organization underneath. PIC-SURE hides this complexity and exposes the different study datasets in a single tabular format. By simplifying the process of data extraction, it allows investigators to focus on the downstream analyses and to facilitate reproducible science.


### More about PIC-SURE
The API is available in two different programming languages, python and R, enabling investigators to query the databases the same way using either language.


PIC-SURE is a larger project from which the R/python PIC-SURE API is only a brick. Among other things, PIC-SURE also offers a graphical user interface that allows researchers to explore variables across multiple studies, filter patients that match criteria, and create cohorts from this interactive exploration.

The R API is actively developed by the Avillach Lab at Harvard Medical School.

PIC-SURE API GitHub repo:
* https://github.com/hms-dbmi/pic-sure-r-adapter-hpds


 -------

## Getting your own user-specific security token

**Before running this notebook, please be sure to review the "Get your security token" documentation, which exists in the [`README.md` file](../README.md). It explains about how to get a security token, which is mandatory to access the databases.**

### Environment set-up

#### Pre-requisites
- R 4.1 or later

#### Install packages

The first step to using the PIC-SURE API is to install the packages needed. The following code installs the PIC-SURE API components from GitHub, specifically the PIC-SURE R Adapter.

**Note that if you are using the dedicated PIC-SURE environment within the *BDC Powered by Seven Bridges (BDC-Seven Bridges)* platform, the necessary packages have already been installed.**

*Troubleshooting: if you encounter an error while installing packages or connecting to a PIC-SURE resource, try restarting your kernel and trying again.*

```{r}
### Uncomment this code if you are not using the PIC-SURE environment in *BDC-Seven Bridges*, or if you do not have all the necessary dependencies installed.
#install.packages("devtools")
```

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
devtools::install_github("hms-dbmi/pic-sure-r-adapter-hpds", ref="query_v3", force=T, quiet=FALSE)
library(dplyr)
```

```{r}
install.packages("stringr")
library(stringr)
```

## Connecting to a PIC-SURE resource
The following is required to get access to data through the PIC-SURE API:
- the platform that should be connected (e.g., BDC Authorized)
- a personalized access token

The following code connects to the *BDC-PIC-SURE* platform using the `picsure::connect` function and references the user-specific token saved as `token.txt`.

If you have not already retrieved your user-specific token, please refer to the "Get your security token" section of the `README.md` file and the `Workspace_setup.ipynb` file.

```{r}
library(picsure)
token_file <- "token.txt"
my_token <- readLines(token_file, warn = FALSE)[1]
session <- picsure::connect(
  platform = picsure::Platform$BDC_AUTHORIZED,
  token    = my_token
)
```

### Getting help with the PIC-SURE API

You can get help with PIC-SURE library functions by using the `?` operator

```{r}
?picsure::searchDictionary
```

For example, the above output lists and briefly defines how to use the `searchDictionary` function.

## Using the PIC-SURE variable dictionary

For the rest of this example notebook, we will use one of the publicly available datasets on PIC-SURE. This dataset is the "Framingham Heart Study: Dataset for Teaching Purposes", which is listed as `tutorial-biolincc_framingham` in the PIC-SURE output above. To find the variables related to this dataset, we will use the `searchPicsure` function to search for a term of interest - in this case, `tutorial`.

```{r}
search_term <- "tutorial"
```

```{r}
my_variables_df <- picsure::searchDictionary(session, search_term, include_values = TRUE) # Search for variables related to 'tutorial'
```

We can now review the variables that were returned and some of the information associated with these variables.

```{r}
nrow(my_variables_df) # How many variables did the search return?
```

```{r}
head(my_variables_df) # Take a peek at some of the variables returned
```

The `conceptPath` column contains the concept paths for specific variables, which are used for building cohorts and selecting participant-level data. Let's take a look at the first ten concept paths in our search results.

```{r}
head(my_variables_df$conceptPath, 10) # What are the first ten concept paths in my search results?
```
We can also view additional information for individual variables by filtering the data dictionary on `conceptPath`. Let's take a look at the details from the first concept path.

```{r}
first_var <- my_variables_df$conceptPath[1] # Save the concept path for the first variable
my_variables_df[my_variables_df$conceptPath == first_var, ] # Show information for that first concept path
```

PIC-SURE integrates clinical and genomic datasets across BDC, including TOPMed and TOPMed-related studies, COVID-19 studies, and BioLINCC studies. Each variable is organized as a concept path that contains information about the study, variable group, and variable. Though the specifics of the concept paths are dependent on the type of study, the overall information included is the same. 

Data Organization in PIC-SURE
---------------------------------------
| Data organization | TOPMed & TOPMed-related studies | BioLINCC & COVID-19 studies |
|-------------------|---------------------------------|-----------------------------|
| General organization | Data organized using the format implemented by the database of Genotypes and Phenotypes (dbGaP). Generally, a given study will have several tables, and those tables have several variables. | Data do not follow dbGaP format; there are no phv or pht accessions. Data are organized in groups of like variables, when available. For example, variables like Age, Gender, and Race could be part of the Demographics variable group. |
| Concept path structure | \phs\pht\phv\variable name\ | \phs\variable name |
| Variable ID | phv corresponding to the variable accession number | Equivalent to variable name | 
| Variable name | Encoded variable name that was used by the original submitters of the data | Encoded variable name that was used by the original submitters of the data |
| Variable description | Description of the variable | Description of the variable, as available |
| Dataset ID | pht corresponding to the trait table accession number | Equivalent to Dataset name | 
| Dataset name | Name of the trait table | Name of a group of like variables, as available | 
| Dataset description | Description of the trait table | Description of a group of like variables, as available |
| Study ID | phs corresponding to the study accession number | phs corresponding to the study accession number |
| Study description | Description of the study from dbGaP | Description of the study from dbGaP |

*Note: The concept paths in PIC-SURE are used for querying. This is called `name` in the data dictionary shown above.*

Now you can try to search for a term on your own. Below is sample code on how to search for the term `sex`. To practice searching the data dictionary, you can change `"sex"` to a term you are interested in. You will see the results displayed as a data frame. Note - the results displayed will show results from all studies you have access to.

```{r}
my_search <- picsure::searchDictionary(session, "sex") # Change "sex" to a term you are interested in
#head(my_search) # Show the variables that match my search result
```


## Using PIC-SURE to build a query and export results
You can also use the PIC-SURE API to build a query and retrieve data. With this functionality, you can filter based on specific variables, add others, and export the data as a data frame into this notebook.

### Building a query with clauses and clause groups
The PIC-SURE R API supports constructing simple and complex queries using individual filter clauses, logical operators (`AND`, `OR`), and nested filter groups.

A **clause** represents a single filtering condition on a variable, such as filtering to specific categorical values or a numeric range. Use `picsure::buildClause()` to create individual clauses.

Multiple clauses can be combined into a **clause group** using `picsure::buildClauseGroup()`. Clause groups allow filters to be joined using either an `AND` or `OR` operator, enabling you to express more complex query logic. Clause groups can also contain other clause groups, making it possible to create nested filtering logic.

Once the filtering logic has been defined, use `picsure::buildQuery()` to assemble the final query. The `phenotypicFilter` argument specifies the clauses or clause groups to include in the query. The `includeConcepts` argument adds variables to the output without applying any additional filtering.

In summary, the general workflow for constructing a query is:

1. Create one or more clauses with `picsure::buildClause()`.
2. Combine clauses into clause groups with `picsure::buildClauseGroup()`.
3. Optionally nest clause groups to create more advanced logic.
4. Build the final query with `picsure::buildQuery()`, specifying additional output variables using `includeConcepts`.
5. Run the query to export the specified participant-level data for a cohort using `picsure::runQuery()`.

There are several clause types that can be used to build a query, which can be specified by the `type` argument in the `picsure::buildClause` function.

| type | functionality |
|--------|-------------------|
| `picsure::PhenotypicFilterType$FILTER` | return input variable; only records that match filter criteria |
| `picsure::PhenotypicFilterType$REQUIRE` | return all variables included; only records that do not contain null values for input variables |
| `picsure::PhenotypicFilterType$ANYRECORD` | return all variables included; only records that contain at least one non-null value for input variables |

### Example with PIC-SURE data
As an example query, let's use the Framingham tutorial dataset to investigate the prevalence of hypertension and distribution of age of current smokers with body mass index greater than 20.

In other words, we want to create a query that filters to study participants that are:
* Current smokers `AND`
* Have BMI greater than 20 `AND`
* Have some measurement for hypertension `AND`
* Have some measurement for age

Let's start by retrieving the variables for the `tutorial-biolincc_framingham` study from our initial search.

```{r}
# Ensure that only Framingham tutorial variables are shown in the data dictionary, which can vary based on individual access
phs_number = "tutorial-biolincc_framingham"
tutorial_df <- my_variables_df[my_variables_df$studyId == phs_number,]
```

### Build a query with a categorical variable - Current smoker
For the first part of our query, we are interested in current smokers. Based on the search for the Framingham tutorial dataset variables, we can save the concept path of the "Current cigarette smoking at exam" variable, which is a categorical variable.

```{r}
smoke_variable_path <- tutorial_df[tutorial_df$description == "Current cigarette smoking at exam", "conceptPath"]
smoke_variable_path # See the concept path
```

We can take a look at the options of values to filter by using the `values` column of the data dictionary.

```{r}
tutorial_df[tutorial_df$conceptPath == smoke_variable_path, "values"]
```

Let's apply a filter on the "Current cigarette smoking at exam" variable to only select participants with "Current smoker." Note that though we are only filtering by one value, you can filter by multiple values by passing a list into `filter()`. Note that though we are only filtering by one value, you can filter by multiple values by passing a list into the `filter` function. 

```{r}
smoke_clause <- picsure::buildClause(smoke_variable_path,
                                      type       = picsure::PhenotypicFilterType$FILTER,
                                      categories = c("Current smoker"))
```

### Build a clause with a continuous variable - BMI
For the second part of our query, we are interested in participants with a body mass index (BMI) greater than 20. We can find the applicable variables using a similar process; by searching for the term of interest in the variables.

```{r}
bmi_variable_path <- tutorial_df[tutorial_df$name == "BMI", "conceptPath"]
bmi_variable_path
```

We can take a look at the minimum and maximum values of the variable using the `min` and `max` columns of the data dictionary.

```{r}
tutorial_df[tutorial_df$conceptPath == bmi_variable_path, c("min", "max")]
```

To apply this filter, we will create a separate clause. Let's apply a filter on the "Body Mass Index, weight in kilograms/height meters squared" variable to select only participants with values greater than 20. Note that while in this example only a `min` is specified, a `max` can also be defined for the filter.

```{r}
bmi_clause <- picsure::buildClause(bmi_variable_path,
                                    type = picsure::PhenotypicFilterType$FILTER,
                                    min  = 20)
```

### Building a clause with `REQUIRE` - Age and Hypertension
In addition to adding the filters, we are interested in participants with age and hypertension measurements. Let's do this for the "Age at exam (years)" and "HYPERTEN" variables.

```{r}
age_variable_path <- tutorial_df[tutorial_df$description == "Age at exam (years)", "conceptPath"]
hyperten_variable_path <- tutorial_df[tutorial_df$name == "HYPERTEN", "conceptPath"]
```

We can create separate clauses for these variables as well using `REQUIRE` to be added to the final query later. Note that `REQUIRE` selects participants that have non-null values for these variables; in other words, selects participants that have some measurement.

Note that, here, we need to create two separate clauses to later be combined with `AND`. If both variables are added to a single clause at once, the clause would add these variables with `OR`.

```{r}
# REQUIRE applies per concept path; a single multi-path clause would OR them.
# To require BOTH age and hypertension, build one REQUIRE clause per path and AND them.
age_require_clause      <- picsure::buildClause(age_variable_path,      type = picsure::PhenotypicFilterType$REQUIRE)
hyperten_require_clause <- picsure::buildClause(hyperten_variable_path, type = picsure::PhenotypicFilterType$REQUIRE)
```

### Building the query
Now we have four clauses that should be put together into a single query. We will create a clause group using `picsure::buildClauseGroup` and specify the operator as `AND`.

```{r}
# Put all clauses together with AND
all_clauses <- picsure::buildClauseGroup(
  list(smoke_clause, bmi_clause, age_require_clause, hyperten_require_clause),
  operator = picsure::GroupOperator$AND
)
```

Once the clause group is created, we can build the final query with `picsure::buildQuery()`.

Bonus: Let's say we are interested in reviewing the data for a common complication of smoking and hypertension, like stroke. We don't want to use this variable to filter our cohort, but we are interested in including it in the export for review. We can pass the concept path for the stroke variable into the `includeConcepts` argument of `picsure::buildQuery()`.

```{r}
stroke_variable_path <- tutorial_df[tutorial_df$name == "STROKE", "conceptPath"]

example_query <- picsure::buildQuery(
  phenotypicFilter = all_clauses,
  includeConcepts  = stroke_variable_path
)
```

### Running the query and exporting participant-level data
The query has been constructed and can now be run. `picsure::runQuery()` with `type = picsure::QueryType$PARTICIPANT` returns a data frame of participant-level data.

In the data dictionary data frame shown previously, each row represented a single concept path or variable. In the query data frame, the concept paths are added as columns with each row representing a participant with data that matches your query.

The data frame will contain some automatically exported concept paths, such as `patient_id`, and the concept paths we added to our query.

```{r}
example_results <- picsure::runQuery(session, example_query, type = picsure::QueryType$PARTICIPANT)
head(example_results)
```

As you can see in the results, there are some instances where study participants may have more than one value for a given variable. For example, this may be the case when a study participants answers questionnaires for multiple visits. 

In the PIC-SURE output, this is shown as values being separated by a tab or `\t` value. These multiple values will need to be accounted for depending on the planned analysis.

With this example, averages of the age and BMI values could be calculated and a new variable "ever smoker" could be created based on whether "current smoker" was ever answered for the CURSMOKE variable. The code below shows this example of how to handle these values.

*Note: Approaches to handling multiple values will differ based on the approach.*

```{r}
# Select rows of interest and rename them
clean_results <- example_results %>% 
  select("\\tutorial-biolincc_framingham\\AGE\\", "\\tutorial-biolincc_framingham\\BMI\\", "\\tutorial-biolincc_framingham\\CURSMOKE\\", "\\tutorial-biolincc_framingham\\HYPERTEN\\", "\\tutorial-biolincc_framingham\\STROKE\\") %>%
  rename(AGE = "\\tutorial-biolincc_framingham\\AGE\\", BMI = "\\tutorial-biolincc_framingham\\BMI\\", CURSMOKE = "\\tutorial-biolincc_framingham\\CURSMOKE\\", HYPERTEN = "\\tutorial-biolincc_framingham\\HYPERTEN\\", STROKE = "\\tutorial-biolincc_framingham\\STROKE\\")
head(clean_results)
```

```{r}
# Function that splits the values and calculates the mean
mean_multiple_values <- function(df_values) {
  sep_values <- str_split(df_values, "\t")
  numeric_values <- as.numeric(sep_values[[1]])
  return(mean(numeric_values))
}

# Apply the function to calculate the means to the AGE and BMI variables
mean_age <- lapply(clean_results$AGE, mean_multiple_values)
mean_bmi <- lapply(clean_results$BMI, mean_multiple_values)
```

```{r}
# Function that flags participants as smoker if they have an answer of "Current smoker"
ever_smoker <- function(smoke_vals){
  sep_smoke_vals <- str_split(smoke_vals, "\t")[[1]]
  if ("Current smoker" %in% sep_smoke_vals) {
    return("Smoker")
  } else {
    return("Non-smoker")
  }
}

# Apply the function to identify smokers to the CURSMOKE variable
ever_smoker <- lapply(clean_results$CURSMOKE, ever_smoker)
```

```{r}
# Take a look at the new results
clean_results$mean_age <- as.numeric(mean_age)
clean_results$mean_bmi <- as.numeric(mean_bmi)
clean_results$ever_smoker <- as.character(ever_smoker)
clean_results %>% select("mean_age", "mean_bmi", "ever_smoker", "HYPERTEN", "STROKE") %>% head()
```