---
title: An R Markdown document converted from "2_TOPMed_DCC_Harmonized_Variables_analysis.ipynb"
output: html_document
---

# Accessing *NHLBI BioData Catalyst® (BDC)* Harmonized variables using R PIC-SURE API

This tutorial notebook will demonstrate how to query and work with the *BDC* cross-studies harmonized variables using R PIC-SURE API. For a more step-by-step introduction to the R PIC-SURE API, see the `1_PICSURE_API_101.ipynb` notebook.

**Before running this notebook, please be sure to review the \"Get your security token\" documentation, which exists in the [`README.md` file](../README.md). It explains about how to get a security token, which is mandatory to access the databases.**

 -------

# Environment set-up

## System requirements
- R 4.1 or later

## Install Packages

**Note that if you are using the dedicated PIC-SURE environment within the *BDC Powered by Seven Bridges (BDC-Seven Bridges)* platform, the necessary packages have already been installed.**

*Troubleshooting: if you encounter an error while installing packages or connecting to a PIC-SURE resource, try restarting your kernel and trying again.*

```{r}
### Uncomment this code if you are not using the PIC-SURE environment in *BDC-Seven Bridges*, or if you do not have all the necessary dependencies installed.
#install.packages(c("devtools", "dplyr", "ggplot2"))
```

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
devtools::install_github("hms-dbmi/pic-sure-r-adapter-hpds", ref="query_v3", force=T, quiet=FALSE)
install.packages("tidyverse")
library(picsure)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Connecting to a PIC-SURE resource

```{r}
library(picsure)
token_file <- "token.txt"
my_token <- readLines(token_file, warn = FALSE)[1]
session <- picsure::connect(platform = picsure::Platform$BDC_AUTHORIZED, token = my_token)
```

## About Harmonized Variables

The data harmonization effort aims to produce "a high quality, lasting resource of publicly available and thoroughly documented harmonized phenotype variables". The TOPMed Data Coordinating Center collaborates with working group members and phenotype experts on this endeavor. To learn more about which variables are available on BioData Catalyst Powered by PIC-SURE, please refer to the [table of TOPMed DCC Variables in the BDC-PIC-SURE documentation](https://pic-sure.gitbook.io/nhlbi-biodata-catalyst-powered-by-pic-sure/appendix/appendix-2-table-of-topmed-dcc-harmonized-variables-in-pic-sure). 

The following phenotypes are included as harmonized variables:

- Key NHLBI phenotypes    
    - Blood cell counts
    - VTE
    - Atherosclerosis-related phenotypes
    - Lipids
    - Blood pressure


- Common covariates
    - Height
    - Weight
    - BMI
    - Smoking status
    - Race/ethnicity

More information about the variable harmonization process is available at https://www.nhlbiwgs.org/sites/default/files/pheno_harmonization_guidelines.pdf

# Exploring the DCC Harmonized Variables in PIC-SURE
## 1. Identifying harmonized variables of interest using faceted search
First, let's explore what variables are available for the TOPMed DCC Harmonized dataset. Since we are interested in variables from one study in particular, we can use the faceted search in PIC-SURE. To see what options are available with faceted search, we can run `session$showAllFacets()`.

```{r}
session$showAllFacets()
```

This returns a data frame with six columns:

| Column | Description |
|---|---|
| `category` | Category identifier (e.g. `dataset_id`) — pass this to `picsure::addFacet`. |
| `Category Display` | Human-readable category label (e.g. `Dataset`). |
| `display` | Facet option's display label. |
| `description` | Facet option's description, if any. |
| `value` | Option identifier — pass this to `picsure::addFacet`. |
| `count` | Number of concepts matching this option. |

To use faceted search, we can build a `facets` object with the category and value(s) we want to use to narrow down search results. In this case, we are interested in the TOPMed DCC Harmonized dataset. To build out this faceted search, we can use the `dataset_id` category with the `DCC Harmonized data set` value with `picsure::addFacet()`.

```{r}
facets <- picsure::facets(session) # Set up facets object
picsure::addFacet(facets, "dataset_id", "DCC Harmonized data set") # Add a facet to filter for the DCC Harmonized data set
facets$view() # We can confirm that the facet was added
```

Now that we've created our facets object, we can use this for search. The facets can be combined with a search term of interest, but since we are interested in all variables from the DCC Harmonized data set, we can use the facet with an empty search.

```{r}
harmonized_df <- picsure::searchDictionary(session, facets = facets) # Search the dictionary using the facet we just set up
head(harmonized_df)
```

Our variable results include subject IDs and 'metadata variables' which contain information about the age of the subject when a certain measure was taken, or the units of a variable. Let's exclude these.

```{r}
# Discarding "subject ID",
# the variables which only indicate age of the subject at which a given harmonized variable was measured,
# and variables which indicate the units of a given harmonized variable
vars_to_remove <- grepl("age_at|SUBJECT_ID|unit_", harmonized_df$conceptPath)
harmonized_df <- harmonized_df[!vars_to_remove, ]
dim(harmonized_df)
```

## 2. Selecting variables and retrieving data from the database
Let's say we are interested in exploring the subset of harmonized variables pertaining to patient demographics. Let's do a search to see which variables have "demographic" as part of the concept path.

```{r}
demographic_df <- harmonized_df[grepl("demographic", harmonized_df$conceptPath), ] # Search for "demographic" in the concept path
head(demographic_df)
vars_of_interest <- demographic_df$conceptPath # Save concept paths
```

## 3. Build and run a PIC-SURE query for participant-level demographics data
Now that we have identified the concept paths related to demographics, we can use these to build a PIC-SURE query. As a reminder, we first build clauses with single variable filters, then clause groups consisting of multiple clauses, and finally construct the query.

Let's build the clause group programmatically by adding all demographics variables as `ANYRECORD` filters that are combined with `OR`.

```{r}
# Build a clause group with separate ANYRECORD demographics clauses
anyof_clauses <- lapply(vars_of_interest, function(p) picsure::buildClause(p, type = picsure::PhenotypicFilterType$ANYRECORD))
demographic_clausegroup <- picsure::buildClauseGroup(anyof_clauses, operator = picsure::GroupOperator$OR)

# Build the query
demographic_query <- picsure::buildQuery(phenotypicFilter = demographic_clausegroup)

# Run the query
demographic_results <- picsure::runQuery(session, demographic_query, type = picsure::QueryType$PARTICIPANT)
head(demographic_results)
```

## 4. Visualizing the output: sex and race across study cohorts
### Male:Female ratio across study cohorts

Below, we wrangle the data to calculate the male/female sex ratio per study cohort and prepare our data for plotting.

```{r}
mf_df <- demographic_results %>%
    select("\\DCC Harmonized data set\\demographic\\annotated_sex_1\\", "\\DCC Harmonized data set\\demographic\\subcohort_1\\")
colnames(mf_df) <- c('sex', 'subcohort')
mf_df <- mf_df %>%
    group_by(subcohort) %>%
    filter(subcohort != '') %>% #filter out empty subcohorts
    summarise(female = sum(sex == 'Female'),
              male = sum(sex == 'Male')) %>%
    mutate(mf_ratio = male / female)

mf_df
```

```{r}
options(repr.plot.width=16, repr.plot.height=12)
mf_df %>%
ggplot(aes(x = subcohort, y = mf_ratio)) + 
    geom_col() + 
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
    labs(x = 'Study Cohort', y = 'Male/Female ratio', title = 'Male/Female ratio across study cohorts')
    
```

### Participant race percentages across study cohorts
Below, we wrangle the data to calculate the percentage of participants in given racial categories per study cohort and prepare out data for plotting.

```{r}
race_df <- demographic_results %>%
    select("\\DCC Harmonized data set\\demographic\\race_us_1\\", "\\DCC Harmonized data set\\demographic\\subcohort_1\\")
colnames(race_df) <- c('race', 'subcohort')
race_df <- race_df %>%
    filter(subcohort != '', race != '') %>% # filter out blanks
    group_by(subcohort, race) %>%
    summarise(n = n())
head(race_df)
```

```{r}
ggplot(race_df, aes(fill=race, y=n, x=subcohort)) + 
    geom_bar(position="fill", stat="identity") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) +
    labs(x = 'Study Cohort', y = 'Race percentage', title = 'Race percentage distribution across study cohorts')
    
```

