---
title: An R Markdown document converted from "4_Genomic_Queries.ipynb"
output: html_document
---

# *NHLBI BioData Catalyst® (BDC)* Powered by PIC-SURE API Use-Case: Querying on Genomic Variables

This tutorial notebook focuses how to use the PIC-SURE API to query genomic and phenotypic data together.

For a more basic introduction to the python PIC-SURE API, see the `1_PICSURE_API_101.ipynb` notebook.
 
**Before running this notebook, please be sure to get a user-specific security token. For more information about how to proceed, see the \"Get your security token\" instructions in the [`README.md` file](../README.md).**

 -------

# Environment set-up

### Pre-requisites
- R 3.4 or later

### Install packages

**Note that if you are using the dedicated PIC-SURE environment within the *BDC Powered by Seven Bridges (BDC-Seven Bridges)* platform, the necessary packages have already been installed.**

*Troubleshooting: if you encounter an error while installing packages or connecting to a PIC-SURE resource, try restarting your kernel and trying again.*

```{r}
### Uncomment this code if you are not using the PIC-SURE environment in *BDC-Seven Bridges*, or if you do not have all the necessary dependencies installed.
#install.packages(c("devtools", "dplyr"))
```

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
devtools::install_github("hms-dbmi/pic-sure-r-adapter-hpds", ref = "main", force = TRUE, quiet = FALSE)
library(dplyr)
```

### Connecting to a PIC-SURE resource

```{r}
token_file <- "token.txt"
token <- scan(token_file, what = "character")

session <- picsure::connect(picsure::Platform$BDC_AUTHORIZED, token)
```

# Example 1: Build a query with genomic and phenotypic variables
## 1.1 Building the query with the PIC-SURE API

We are going to create a new query request from the PIC-SURE resource that was specified above.  For this example, we will limit the query to a single dataset (the TOPMed DCC Harmonized dataset) by filtering on sex, body mass index, and two genomic filters. 

For more information about the TOPMed DCC Harmonized dataset, please refer to the [`2_TOPMed_DCC_Harmonized_Variables_analysis.ipynb` notebook](./2_TOPMed_DCC_Harmonized_Variables_analysis.ipynb).

### Search for "sex" in TOPMed DCC Harmonized dataset

We can search for variables within a specific dataset using facets. Let's set up the facets object to search only within the TOPMed DCC Harmonized data set.

```{r}
facets <- picsure::facets(session) # Set up facets object
facets <- picsure::addFacet(facets, "dataset_id", "DCC Harmonized data set") # Add a facet to filter for the DCC Harmonized data set
facets$view() # We can confirm that the facet was added
```

Now that we've created our facets object, we can use this for search. We will use this to search for "sex".

```{r}
sex_dataframe <- picsure::searchDictionary(session, term = "sex", facets = facets) # Search the dictionary using the facet we just set up
head(sex_dataframe)
```

Now that we have found variables related to our search, we can apply a filter to select Females.

```{r}
sex_var <- sex_dataframe %>% filter(name == "annotated_sex_1")
sex_var
```

Let's examine the values within this variable.

```{r}
sex_var %>% pull(values)
```

We can create a clause to filter to only those participants with a `Female` sex value.

```{r}
sex_clause <- picsure::buildClause(sex_var$conceptPath, picsure::PhenotypicFilterType$FILTER, categories = "Female")
```

### Search for "body mass index" in TOPMed DCC Harmonized dataset

For this example, we are only interested in obese participants (BMI >= 30). We can use the same facets object to search for "body mass index" in the TOPMed DCC Harmonized dataset.

```{r}
# Search the TOPMed Harmonized dataset for a BMI variable
bmi_dataframe <- picsure::searchDictionary(session, term = "body mass index", facets = facets) # Search the dictionary using the facet we just set up
head(bmi_dataframe)
```

Now we can create a new clause to filter to participants with BMI greater than 30.

```{r}
bmi_clause <- picsure::buildClause(bmi_dataframe$conceptPath, picsure::PhenotypicFilterType$FILTER, min = 30)
```

## 1.2 Building the query with the PIC-SURE API: Genomic Filters

There are several fields that are available to add as part of a genomic filter. The two main fields are `Gene_with_variant` and `Variant_consequence_calculated`. We can search across these fields using `searchGenomicValues()`.

```{r}
# Peak at a list of genes with variant
gene_search <- picsure::searchGenomicValues(session, "Gene_with_variant")
head(gene_search)
```
```{r}
# Peak at a list of variant consequences and severities
consequence_search <- picsure::genomicConsequences()
head(consequence_search)
```

The fat mass and obesity related gene, or *FTO* gene, [has been linked to obesity and other diseases](https://www.ncbi.nlm.nih.gov/gene/79068). Let's search for this gene.

```{r}
# Validate that FTO gene is an option in gene search
gene_search <- picsure::searchGenomicValues(session, "Gene_with_variant", query = "FTO")
gene_search
```

Now we can add the filters to the query.

```{r}
# Build clauses for the FTO gene and high severity consequences
fto_gene_clause <- picsure::buildGenomicFilter(picsure::GenomicFilterKey$GENE_WITH_VARIANT, values = "FTO")
high_severity_clause <- picsure::buildGenomicFilter(picsure::GenomicFilterKey$VARIANT_SEVERITY, values = picsure::VariantSeverity$HIGH)
```


## 1.3 Construct query and retrieve data
Now let's put together the query clauses that filter to participants with:

* Sex = Female
* BMI > 30
* Participants have a highly severe variant on FTO

We will run a count query to find the number of matching participants. This is a great way to preview how many participants match your query criteria without extracting all of the data yet.

```{r}
pheno_group <- picsure::buildClauseGroup(list(sex_clause, bmi_clause), operator = picsure::GroupOperator$AND)

full_query <- picsure::buildQuery(phenotypicFilter = pheno_group, genomicFilters = list(fto_gene_clause, high_severity_clause))
count_result <- picsure::runQuery(session, full_query, type = picsure::QueryType$COUNT)
count_result
```

We will retrieve our results in a dataframe. Note that since we only added sex and BMI to the query, these are the only phenotypic variables returned.

```{r}
full_results <- picsure::runQuery(session, full_query, type = "participant")
head(full_results)
```


# Example 2: Use case with *SERPINA1* gene variants as a risk factor for COPD

In this example, we will create a query to explore the relationship between the COPD phenotype and variants in the SERPINA1 gene. Variations of the SERPINA1 gene have been found to be a strong risk factor for COPD, which you can read more about [here](https://pubmed.ncbi.nlm.nih.gov/31661293/).

To explore this relationship, we will narrow the cohort down to participants that meet the following criteria:
* participated in the COPDgene study
* have had COPD
* have a missense variant of the *SERPINA1* gene

## 2.1 Create query
Let's start by creating a facets object to search for variables within the COPDgene study (phs000179).

```{r}
copd_facets <- picsure::facets(session) # Set up facets object
copd_facets <- picsure::addFacet(copd_facets, "dataset_id", "phs000179")
copd_facets$view() # We can confirm that the facet was added
```

### Criteria 1: Participants who have had COPD
Let's search for COPDGene variables that are related to COPD.

```{r}
# Search the COPDGene study for COPD variable
copd_dataframe <- picsure::searchDictionary(session, term = "copd", facets = copd_facets) # Search the dictionary using the facet we just set up
head(copd_dataframe)
```

The variable with the description "COPD: have you ever had COPD" seems the most promising. Let's copy that variable's HPDS path to use in our query. Also note what values are present for this variable- we will filter to participants with value "Yes".

```{r}
copd_path <- '\\phs000179\\pht002239\\phv00159731\\COPD\\'
```

```{r}
copd_clause <- picsure::buildClause(copd_path, picsure::PhenotypicFilterType$FILTER, categories = "Yes")
```

### Criteria 2: Participants with a missense variant on *SERPINA1*

A filter for the *SERPINA1* gene can be added in a way similar to the first query shown above.

```{r}
serpina_gene_clause <- picsure::buildGenomicFilter(picsure::GenomicFilterKey$GENE_WITH_VARIANT, values = "SERPINA1")
```

We can take a look at the variant consequences available for filtering to confirm that missense variant is a filtering option.

```{r}
missense_variant_clause <- picsure::buildGenomicFilter(picsure::GenomicFilterKey$VARIANT_CONSEQUENCE_CALCULATED, values = "missense_variant")
```

COPDGene, like many studies in BDC, has associated participant genomic data in the TOPMed version of the study. In order to map to these genomic files, we will need the sample identifier from the TOPMed study of COPDGene, which has an accession number of phs000951.

```{r}
copd_topmed_facets <- picsure::facets(session)
copd_topmed_facets <- picsure::addFacet(copd_topmed_facets, "dataset_id", "phs000951")
sample_id_search <- picsure::searchDictionary(session, term = "sample id", facets = copd_topmed_facets)
head(sample_id_search)
```

To add these variables to the query, we can create a new clause that includes these as concepts in the export once we build the query. For now, we will save them for later.

```{r}
sample_vars <- sample_id_search %>% filter(display == "SAMPLE_ID") %>% pull(conceptPath)
sample_vars
```

### 2.2 Get Results
We can construct the query to retrieve counts and the full data. Let's make the query.

```{r}
full_query <- picsure::buildQuery(phenotypicFilter = copd_clause, genomicFilters = list(serpina_gene_clause, missense_variant_clause), includeConcepts = sample_vars)
full_query_count <- picsure::runQuery(session, full_query, type = picsure::QueryType$COUNT)
full_query_count
```

```{r}
full_query_results <- picsure::runQuery(session, full_query, type = picsure::QueryType$PARTICIPANT)
head(full_query_results)
```

The sample identifier column can be used to identify genomic data files from the associated TOPMed study for analysis. The following code can be used to get a list of the sample IDs associated with your cohort. Note that there are multiple samples associated with a single participant. In these cases, the IDs are separated by a tab or `\t` character.
