---
title: An R Markdown document converted from "6_Sickle_Cell.ipynb"
output: html_document
---

# *NHLBI BioData Catalyst® (BDC) Powered by PIC-SURE* API use-case: quick analysis on Hematopoietic Cell Transplant for Sickle Cell Disease (HCT for SCD) data

This is a tutorial notebook aimed to get the user quickly up and running with the R PIC-SURE API. It covers the main functionalities of the API.

 -------   

# Getting your own user-specific security token

**Before running this notebook, please be sure to review the "Get your security token" documentation, which exists in the [`README.md` file](../README.md). It explains about how to get a security token, which is mandatory to access the databases.**

# Environment set-up

### Pre-requisites
- R 3.4 or later

### Install Packages

**Note that if you are using the dedicated PIC-SURE environment within the *BDC Powered by Seven Bridges (BDC-Seven Bridges)* platform, the necessary packages have already been installed.**

*Troubleshooting: if you encounter an error while installing packages or connecting to a PIC-SURE resource, try restarting your kernel and trying again.*

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
```

```{r}
### Uncomment this code if you are not using the PIC-SURE environment in *BDC-Seven Bridges*, or if you do not have all the necessary dependencies installed.
#install.packages(c("devtools"))
```

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
devtools::install_github("hms-dbmi/pic-sure-r-adapter-hpds", ref="main", force=T, quiet=FALSE)
library(dplyr)
library(picsure)
```

## Connecting to a PIC-SURE resource

The following is required to get access to data through the PIC-SURE API: 
- Network URL
- Resource id
- User-specific security token

If you have not already retrieved your user-specific token, please refer to the "Get your security token" section of the `README.md` file.

```{r}
PICSURE_network_URL = "https://picsure.biodatacatalyst.nhlbi.nih.gov/picsure"
token_file <- "token.txt"
token <- scan(token_file, what = "character")
session <- picsure::bdc.initializeSession(PICSURE_network_URL, token)
session <- picsure::bdc.setResource(session = session,  resourceName = "AUTH")
```

# Sickle Cell Example

Let's search the data dictionary in PIC-SURE to find all the variables in the `HCT for SCD` study, which is associated to PHS study number `phs002385`. 

```{r}
df <- picsure::bdc.searchPicsure(session, 'phs002385')
```

```{r}
head(df)
```

### Building a query: investigating male patients with avascular necrosis who received their transplant after 1999

Let's say we are interested in the age at which patients from the following cohort received their transplant:
* males
* patients with avascular necrosis
* patients who received their transplant after 1999

We will use regular expressions to search the variable descriptions within the HCT for SCD study to find these variables.

```{r}
variables_of_interest <- df[grepl('Sex|Avascular necrosis|Year of transplant$', df$var_description),]

sex_var <- df[grepl('Sex', df$var_description),'name']
avascular_necrosis_var <- df[grepl('Avascular necrosis', df$var_description),'name']
transplant_yr_var <- df[grepl('Year of transplant$', df$var_description),'name']

variables_of_interest
```

Now we can create a new query using the `name` associated with our variables of interest and apply our filters to retrieve the cohort of interest.

```{r}
myquery <- picsure::bdc.newQuery(session) # Initialize a new query

# Filter to Males
myquery <- picsure::bdc.addClause(query = myquery, 
                                  keys = sex_var,
                                  type = 'FILTER',
                                  categories = list('Male'))

# Filter to patients with Avascular Necrosis
myquery <- picsure::bdc.addClause(query = myquery, 
                                  keys = avascular_necrosis_var,
                                  type = 'FILTER',
                                  categories = list('Yes'))
                                  

# Filter to patients with year of transplant after 1999
myquery <- picsure::bdc.addClause(query = myquery, 
                                  keys = transplant_yr_var,
                                  type = 'FILTER',
                                  min = 1999, max = 2019)
```

We are also interested in the patients' age at which they received their transplant. However, we do not want to filter our cohort based on these values. We can use the `SELECT` query type to add this variable to the query without filtering.

Using this cohort, we can add the variable of interest: "Patient age at transplant, years"

```{r}
age_transplant_var <- df[grepl('age at transplant, years$', df$var_description),'name']
myquery <- picsure::bdc.addClause(query = myquery, 
                                  keys = age_transplant_var,
                                  type = 'SELECT')
```

## Retrieving the data

Now that the query object is built, we will use the `query.run` function to retrieve the data corresponding to our query

```{r}
results <-  picsure::runQuery(myquery)
head(results)
```

Once the data has been retrieved as a dataframe, you can use R functions to conduct analyses and create visualizations, such as this:

```{r}
hist(results[,5], breaks = 10, 
     xlab = 'Age received transplant, yrs old', 
     main = "Histogram of the ages at which males with Avascular Necrosis \nreceived their transplant in the HCT for SCD study")
```

