---
title: An R Markdown document converted from "6_Sickle_Cell.ipynb"
output: html_document
---

# *NHLBI BioData Catalyst® (BDC) Powered by PIC-SURE* API use-case: quick analysis on Hematopoietic Cell Transplant for Sickle Cell Disease (HCT for SCD) data

This is a tutorial notebook aimed to get the user quickly up and running with the R PIC-SURE API. It covers the main functionalities of the API.

 -------   

# Getting your own user-specific security token

**Before running this notebook, please be sure to review the "Get your security token" documentation, which exists in the [`README.md` file](../README.md). It explains about how to get a security token, which is mandatory to access the databases.**

# Environment set-up

### Pre-requisites
- R 4.1 or later

### Install Packages

**Note that if you are using the dedicated PIC-SURE environment within the *BDC Powered by Seven Bridges (BDC-Seven Bridges)* platform, the necessary packages have already been installed.**

*Troubleshooting: if you encounter an error while installing packages or connecting to a PIC-SURE resource, try restarting your kernel and trying again.*


```{r}
### Uncomment this code if you are not using the PIC-SURE environment in *BDC-Seven Bridges*, or if you do not have all the necessary dependencies installed.
#install.packages(c("devtools"))
```

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
devtools::install_github("hms-dbmi/pic-sure-r-adapter-hpds", ref="query_v3", force=T, quiet=FALSE)
library(picsure)
library(dplyr)
```

## Connecting to a PIC-SURE resource

The following is required to get access to data through the PIC-SURE API: 
- A valid user-specific security token

If you have not already retrieved your user-specific token, please refer to the "Get your security token" section of the `README.md` file.

```{r}
token_file <- "token.txt"
my_token <- readLines(token_file, warn = FALSE)[1]
session <- picsure::connect(platform = picsure::Platform$BDC_AUTHORIZED, token = my_token)
```

# Sickle Cell Example

Let's search the data dictionary in PIC-SURE to find all the variables in the `HCT for SCD` study, which is associated to PHS study number `phs002385`. 

```{r}
facets <- picsure::facets(session) # Set up facets object
picsure::addFacet(facets, "dataset_id", "phs002385") # Add a facet to filter for the HCT for SCD study

df <- picsure::searchDictionary(session, facets = facets)
print(dim(df))
head(df)
```

### Building a query: investigating male patients with avascular necrosis who received their transplant after 1999

Let's say we are interested in the age at which patients from the following cohort received their transplant:
* males
* patients with avascular necrosis
* patients who received their transplant after 1999

We will use regular expressions to search the variable descriptions within the HCT for SCD study to find these variables.

```{r}
variables_of_interest <- df[grepl('Sex|Avascular necrosis|Year of transplant$', df$description),]

sex_var                <- df[grepl("Sex", df$description), "conceptPath"][1]
avascular_necrosis_var <- df[grepl("Avascular necrosis", df$description), "conceptPath"][1]
transplant_yr_var      <- df[grepl("Year of transplant$", df$description), "conceptPath"][1]

variables_of_interest
```

Now we can create a new query using the `conceptPath` associated with our variables of interest and apply our filters to retrieve the cohort of interest.

```{r}
# age at transplant — included as an output column only (not a row filter)
age_transplant_var <- df[grepl("age at transplant, years$", df$description), "conceptPath"][1]

# Build filter clauses
sex_clause       <- picsure::buildClause(sex_var, type = picsure::PhenotypicFilterType$FILTER, categories = c("Male"))
necrosis_clause  <- picsure::buildClause(avascular_necrosis_var, type = picsure::PhenotypicFilterType$FILTER, categories = c("Yes"))
transplant_clause <- picsure::buildClause(transplant_yr_var, type = picsure::PhenotypicFilterType$FILTER, min = 1999, max = 2019)

# Combine filters with AND logic
filter_tree <- picsure::buildClauseGroup(list(sex_clause, necrosis_clause, transplant_clause), operator = picsure::GroupOperator$AND)

# Build the query; age_transplant_var is requested as an output column via includeConcepts
myquery <- picsure::buildQuery(phenotypicFilter = filter_tree, includeConcepts = c(age_transplant_var))
```

We are also interested in the patients' age at which they received their transplant. However, we do not want to filter our cohort based on these values. We use `includeConcepts` in `buildQuery` to request this variable as an output column without applying it as a filter criterion.

## Retrieving the data

Now that the query object is built, we will use the `runQuery` function to retrieve the data corresponding to our query

```{r}
results <- picsure::runQuery(session, myquery, type = picsure::QueryType$PARTICIPANT)
head(results)
```

Once the data has been retrieved as a dataframe, you can use R functions to conduct analyses and create visualizations, such as this:

```{r}
hist(results[,2], breaks = 10, 
     xlab = 'Age received transplant, yrs old', 
     main = "Histogram of the ages at which males with Avascular Necrosis \nreceived their transplant in the HCT for SCD study")
```

