---
title: An R Markdown document converted from "7_Harmonization_with_PICSURE.ipynb"
output: html_document
---

# Harmonization across studies with *NHLBI BioData Catalyst® (BDC) Powered by PIC-SURE*

This tutorial notebook will demonstrate how to query and work with the *BDC* studies, particularly cross-study harmonization. For a more step-by-step introduction to the python PIC-SURE API, see the `1_PICSURE_API_101.ipynb` notebook.

**Before running this notebook, please be sure to review the \"Get your security token\" documentation, which exists in the [`README.md` file](../README.md). It explains about how to get a security token, which is mandatory to access the databases.**

-----

# Environment set-up

### System Requirements
R >= 4.1

### Install packages

**Note that if you are using the dedicated PIC-SURE environment within the *BDC Powered by Seven Bridges (BDC-Seven Bridges)* platform, the necessary packages have already been installed.**

*Troubleshooting: if you encounter an error while installing packages or connecting to a PIC-SURE resource, try restarting your kernel and trying again.*

```{r}
### Uncomment this code if you are not using the PIC-SURE environment in *BDC-Seven Bridges*, or if you do not have all the necessary dependencies installed.
#install.packages(c("devtools", "dplyr", "ggplot2"))
```

```{r}
Sys.setenv(TAR = "/bin/tar")
options(unzip = "internal")
devtools::install_github("hms-dbmi/pic-sure-r-adapter-hpds", ref="query_v3", force=T, quiet=FALSE)
library(picsure)
library(dplyr)
library(ggplot2)
library(stringr)
```

```{r}
install.packages("BiocManager", verbose=TRUE)
library(BiocManager)
BiocManager::install("limma")
library(limma)
```

## Connecting to a PIC-SURE Network

```{r}
token_file <- "token.txt"
my_token <- readLines(token_file, warn = FALSE)[1]
session <- picsure::connect(platform = picsure::Platform$BDC_AUTHORIZED, token = my_token)
```

-----

## Harmonizing variables with PIC-SURE
One of the key challenges to conducting analyses with several studies is ensuring correct data harmonization, or combining of data from different sources. There are many harmonization techniques, but this notebook will demonstrate an approach to finding and extracting similar variables from different studies in PIC-SURE. Two examples of this will be shown:
1. Retrieving variables for *sex and gender* across studies with BMI
2. Harmonizing the *"orthopnea"* and *"pneumonia"* variables across studies


*For more information about the TOPMed DCC Harmonized Data Set in PIC-SURE, please refer to the [`2_TOPMed_DCC_Harmonized_Variables_analysis.ipynb`](https://github.com/hms-dbmi/Access-to-Data-using-PIC-SURE-API/blob/master/NHLBI_BioData_Catalyst/python/2_TOPMed_DCC_Harmonized_Variables_analysis.ipynb) notebook*

-----

## Sex and gender variables across studies
<font color='darkgreen'>**Goal: Create harmonized variables for sex and BMI which combine data from multiple studies**</font> 

These variables are labeled differently for each of these studies. For example, some use the keyword `sex` while others use `gender`. To account for these differences, we need to develop a way to search for multiple keywords at once.

Let's start by searching for `sex` and apply faceted search for ECLIPSE (Evaluation of COPD Longitudinally to Identify Predictive Surrogate Endpoints) (phs001252) and COPDGene (Genetic Epidemiology of COPD) (phs000179) as our studies of interest.

```{r}
facets <- picsure::facets(session)
picsure::addFacet(facets, "dataset_id", c("phs000179", "phs001252"))
sex_df <- picsure::searchDictionary(session, "sex", facets = facets)
print(dim(sex_df))
head(sex_df)
```

### Which sex variable in ECLIPSE should we use?
We can see that there are many sex variables in ECLIPSE originating from various data tables. We will examine the data associated with these variables to determine which one to use for our analysis.

```{r}
# which sex variables are part of ECLIPSE?
# create a subset of only ECLIPSE sex vars
eclipse_sex_df <- sex_df %>%
    filter(grepl('phs001252', studyId)) %>%
    select(studyId, name, description, values, conceptPath)
eclipse_sex_df
```

We can see that there is a variable called `BICAT` or "Biomarker". This variable was selected because one of the values associated with this variable is "Sex hormone binding globulin" and was returned with our search. However, this is not a sex variable and we should remove this.

```{r}
eclipse_sex_df <- eclipse_sex_df %>% filter(name != "BICAT")
```

Many studies may have multiple variables with similar names and descriptions that come from different data tables. This will vary from study to study based on how the original study was conducted and organized.

We will approach this complication by examining the data associated with each sex variable in ECLIPSE and determining which would be the best fit for our analysis.

Because we want to examine the patient level data, we will create a query using all sex variables in ECLIPSE. We will add these variables to our query using `picsure::PhenotypicFilterType$ANYRECORD` clauses combined with `picsure::GroupOperator$OR`, as we are interested in all observations with a value for any of our chosen variables.

```{r}
# Build and run query for all ECLIPSE sex variables
anyof_clauses <- lapply(eclipse_sex_df$conceptPath, function(p) picsure::buildClause(p, type = picsure::PhenotypicFilterType$ANYRECORD))
anyof_group <- picsure::buildClauseGroup(anyof_clauses, operator = picsure::GroupOperator$OR)
eclipse_sex_query <- picsure::buildQuery(phenotypicFilter = anyof_group)
eclipse_sex_results <- picsure::runQuery(session, eclipse_sex_query, type = picsure::QueryType$PARTICIPANT)
head(eclipse_sex_results)
```

We used `picsure::runQuery()` with `type = picsure::QueryType$PARTICIPANT` to organize our participant level results in a data frame that is easy to work with. By previewing the resulting data frame, we can see that not all of the sex variables are complete for our participant subset.

Let's see which sex variables are the most complete for our dataset by counting the number of empty values in each one.

```{r}
# count the number of empty values in each concept column
concept_cols <- setdiff(colnames(eclipse_sex_results), 'patient_id')
complete_df <- as.data.frame(colSums(eclipse_sex_results[, concept_cols, drop = FALSE] == '', na.rm = TRUE))
colnames(complete_df) <- 'n_empty'
complete_df
```

We can see that there are a few variables with 0 empty values, we would like to focus on these.

```{r}
# focus on those variables which do not have any empty values
non_empty_cols <- complete_df %>% filter(n_empty == 0) %>% rownames
filtered_eclipse_sex_results <- eclipse_sex_results %>% select(all_of(c('patient_id', non_empty_cols)))
head(filtered_eclipse_sex_results)
```

Now we have filtered our dataset to only include sex variables in ECLIPSE which have complete data. Let's check to see if the values in each variable (column) differ, or if they all have the same data.

```{r}
# in this example, we notice that all our filtered variables have equivalent data for each patient
for (col in setdiff(colnames(filtered_eclipse_sex_results), 'patient_id')) {
    cat(col, '\n')
    print(unique(filtered_eclipse_sex_results[[col]]))
}
```

We can see that all the values in the filtered, complete ECLIPSE sex variables are the same, so we will simply select the first variable to use in our analysis.

We can look up a variable in the search result data frame to learn more about it.

```{r}
eclipse_sex_var <- colnames(filtered_eclipse_sex_results)[2]
sex_df[sex_df$conceptPath == eclipse_sex_var, ]
```

### Which sex variable in COPDGene should we use?
We will then repeat the same approach to select a sex variable from the COPDGene study.

```{r}
# which sex variables are part of COPDGene?
# create a subset of only COPDGene sex vars
copdgene_sex_df <- sex_df %>%
    filter(grepl('phs000179', studyId)) %>%
    select(studyId, name, description, values, conceptPath)
copdgene_sex_df
```

```{r}
# Remove the non-sex variables
copdgene_sex_df <- copdgene_sex_df %>% filter(!name %in% c("phv00569257", "phv00569446"))
```

```{r}
# Build and run query for all COPDGene sex variables
copdgene_paths <- copdgene_sex_df$conceptPath
anyof_clauses <- lapply(copdgene_paths, function(p) picsure::buildClause(p, type = picsure::PhenotypicFilterType$ANYRECORD))
anyof_group <- picsure::buildClauseGroup(anyof_clauses, operator = picsure::GroupOperator$OR)
copdgene_sex_query <- picsure::buildQuery(phenotypicFilter = anyof_group, includeConcepts = copdgene_paths)
copdgene_sex_results <- picsure::runQuery(session, copdgene_sex_query, type = picsure::QueryType$PARTICIPANT)
head(copdgene_sex_results)
```

```{r}
# count the number of empty values in each concept column
copdgene_concept_cols <- setdiff(colnames(copdgene_sex_results), 'patient_id')
colSums(copdgene_sex_results[, copdgene_concept_cols, drop = FALSE] == '', na.rm = TRUE)
```

```{r}
# check that all our variables have equivalent data for each patient
for (col in setdiff(colnames(copdgene_sex_results), 'patient_id')) {
    cat(col, '\n')
    print(unique(copdgene_sex_results[[col]]))
}
```

```{r}
# Save the first sex var from COPDGene
copdgene_sex_var <- colnames(copdgene_sex_results)[2]
sex_df[sex_df$conceptPath == copdgene_sex_var, ]
```

### Which BMI variables should we use?

We will follow the approach outlined above to first search our data dictionary for all variables containing the terms 'bmi'. 

```{r}
bmi_df <- picsure::searchDictionary(session, "bmi", facets = facets)
bmi_df
```

By examining metadata like the variable name and description, we are able to determine which variables we are interested in:
- BMI from COPDGene (\phs000179\pht002239\phv00159593\BMI\)
- Body mass index (kg/m2) from the bodycomp table for ECLIPSE (\phs001252\pht006373\phv00293610\VSBMI\) 

We will save the HPDS_PATHs associated to the BMI variables of interest so we can add it to our query later.

```{r}
# save HPDS_PATHs of interest
copdgene_bmi_var = "\\phs000179\\pht002239\\phv00159593\\BMI\\"
eclipse_bmi_var = "\\phs001252\\pht006373\\phv00293610\\VSBMI\\"
```

### Query PIC-SURE for selected variables
We build a query from the selected concept paths and run it against PIC-SURE. As a reminder, these are the paths associated to the sex and BMI variables from ECLIPSE and COPDGene. Our goal is to build a query that returns participants with sex *and* BMI information for either COPDGene *or* ECLIPSE.

Based on this, here is our plan to build a PIC-SURE query:
- ECLIPSE participants with any record for sex *and* BMI

or

- COPDGene participants with any record for sex *and* BMI

We use `picsure::PhenotypicFilterType$ANYRECORD` clauses combined with `picsure::GroupOperator$AND` because we are interested in participants with information for both sex and BMI. Then, these can be combined into a clause group using `picsure::GroupOperator$OR`. If you are interested in learning more about the other query methods available, see the `1_PICSURE_API_101.ipynb` notebook.

```{r}
eclipse_group <- picsure::buildClauseGroup(
    lapply(c(eclipse_sex_var, eclipse_bmi_var), function(p) picsure::buildClause(p, type = picsure::PhenotypicFilterType$ANYRECORD)),
    operator = picsure::GroupOperator$AND
)
copdgene_group <- picsure::buildClauseGroup(
    lapply(c(copdgene_sex_var, copdgene_bmi_var), function(p) picsure::buildClause(p, type = picsure::PhenotypicFilterType$ANYRECORD)),
    operator = picsure::GroupOperator$AND
)

full_group <- picsure::buildClauseGroup(list(eclipse_group, copdgene_group), operator = picsure::GroupOperator$OR)

combined_query <- picsure::buildQuery(phenotypicFilter = full_group)
combined_results <- picsure::runQuery(session, combined_query, type = picsure::QueryType$PARTICIPANT)
head(combined_results)
```

### Clean and wrangle query result dataframe
You may notice that the column names are not the most informative, we will rename them below.

```{r}
# rename concept-path columns to readable labels (order-independent)
combined_results <- combined_results %>%
    rename(
        eclipse_sex  = !!eclipse_sex_var,
        eclipse_bmi  = !!eclipse_bmi_var,
        copdgene_sex = !!copdgene_sex_var,
        copdgene_bmi = !!copdgene_bmi_var
    )
```

We will also create a 'study' column to easier identify from which study a participant's data originated from.

```{r}
# label participant with study
combined_results <- combined_results %>%
    mutate(study = ifelse(is.na(copdgene_sex) | copdgene_sex == '', 'ECLIPSE', 'COPDGENE'))
head(combined_results)
```

Finally, we will harmonize the sex and bmi variables by combining them into one single 'sex' column.

Sex and BMI are reported in the same way in both COPDGene and ECLIPSE, so we do not have to transform any of the participant level data.

```{r}
# combine sex variables
combined_results <- combined_results %>%
    mutate(sex = ifelse(is.na(copdgene_sex) | copdgene_sex == '', eclipse_sex, copdgene_sex)) %>%
    select(-copdgene_sex, -eclipse_sex)
head(combined_results)
```
```{r}
# combine bmi variables
combined_results <- combined_results %>%
    mutate(bmi = ifelse(is.na(copdgene_bmi) | copdgene_bmi == '', eclipse_bmi, copdgene_bmi)) %>%
    select(-copdgene_bmi, -eclipse_bmi)
head(combined_results)
```

### Analysis and visualization

In this example, we are interested in comparing mean BMI across different studies. We will calculate the mean for each group (ECLIPSE males, ECLIPSE females, COPDGene males, COPDGene females) and display the results as a table and a bar chart.

In some instances, a single study participant may have multiple values for a given variable. When this is the case, the values are separated by a "tab", or `\t` character. We can do some cleaning to separate these values - let's take the mean BMI for each individual with multiple values.

```{r}
get_mean_from_multiple_values <- function(x) {
    y <- as.numeric(mean(as.numeric(unlist(str_split(x, '\t'))), na.rm = TRUE))
    return(y)
}
```

```{r}
combined_results$bmi <- sapply(combined_results$bmi, get_mean_from_multiple_values)

plotdf <- combined_results %>% select(sex, study, bmi)
head(plotdf)
```
```{r}
# remove unneccessary columns and summarize
plotdf = combined_results %>% 
    select(sex, study, bmi) %>%
    group_by(sex, study) %>%
    summarise(mean = mean(bmi, na.rm = TRUE))

plotdf
```
```{r}
plotdf %>%
ggplot(aes(x = sex, y = mean, fill = study)) +
    geom_col(position="dodge") +
    theme_bw()
```


## Orthopnea variables across studies
<font color='darkgreen'>**Goal: Create harmonized variables for orthopnea and pneumonia which combine data from multiple studies, and compare participant distributions across these phenotypes**</font>

In this example, we will be harmonizing variables related to `orthopnea`. Orthopnea is shortness of breath that occurs when individuals lie flat. Because of this, people with orthopnea have to sleep propped up in bed or sitting in a chair. You can read more about this condition [here](https://www.sleepfoundation.org/sleep-apnea/orthopnea).

Let's start by doing a search in PIC-SURE for concept paths containing `orthopnea` from these studies:
- FHS (phs000007)
- MESA (phs000209)
- CHS (phs000287)

### Search and find orthopnea variables

```{r}
facets <- picsure::facets(session)
picsure::addFacet(facets, "dataset_id", c("phs000007", "phs000209", "phs000287"))
orthopnea_dataframe <- picsure::searchDictionary(session, "orthopnea", facets = facets)
print(dim(orthopnea_dataframe))
head(orthopnea_dataframe)
```

As shown in the resulting dataframe, orthopnea is often recorded as something like `sleep on two or more pillows to help you breathe?`

Let's do a search for `pillows` in PIC-SURE to explore those concept paths.

```{r}
pillows_dataframe <- picsure::searchDictionary(session, "pillows", facets = facets)
head(pillows_dataframe)
```

As shown in the resulting dataframe, there are some concept paths related to orthopnea that were not captured by the previous `orthopnea` search, such as `SLEEP ON TWO OR MORE PILLOWS TO HELP BREATHE` from the MESA study.

This is problematic for researchers studying these types of variables, since the concept paths of interest may not always be captured by a single search term. To account for this, researchers must conduct exploratory searches to determine potential search terms.

```{r}
pillows_dataframe[pillows_dataframe$name == "phv00099056", ]
```
```{r}
orthopnea_variables_of_interest <- orthopnea_dataframe[orthopnea_dataframe$name == 'phv00021018', ]
pillows_variables_of_interest <- pillows_dataframe[pillows_dataframe$name %in% c('phv00099056', 'phv00087203'), ]

# save variable paths for querying data
orthopnea_variable_paths_of_interest <- c(orthopnea_variables_of_interest$conceptPath, pillows_variables_of_interest$conceptPath)
```

### Search and find pneumonia variables
As part of our research, let's say we are interested in exploring the relationship between pneumonia and orthopnea. Let's save concept paths related to `pneumonia` as well. 

```{r}
# search for 'pneumonia'
harmonized_df <- picsure::searchDictionary(session, "pneumonia", facets = facets)
print(dim(harmonized_df))
harmonized_df %>% select(name, description, values, studyId)
```

After browsing the available variables, we decided to choose the following:

```{r}
pneumonia_variables_of_interest <- harmonized_df %>%
    filter(grepl('phv00021300|phv00087211|phv00099077', conceptPath))

# save variable paths for querying data
pneumonia_variable_paths_of_interest <- pneumonia_variables_of_interest$conceptPath

pneumonia_variables_of_interest %>% select(values, studyId, description, name)
```

### Query PIC-SURE for selected variables
Our goal is to retrieve participant-level data across several studies for orthopnea and pneumonia. With this in mind, here is our plan to create the PIC-SURE query:
- participants with any record for orthopnea in FHS *or* MESA *or* CHS

and

- participants with any record for pneumonia in FHS *or* MESA *or* CHS

We use `picsure::PhenotypicFilterType$ANYRECORD` clauses combined with `picsure::GroupOperator$OR` because we are interested in participants with information across studies. Then, these can be combined into a clause group using `picsure::GroupOperator$AND`. If you are interested in learning more about the other query methods available, see the `1_PICSURE_API_101.ipynb` notebook.

```{r}
# Build and run query for orthopnea and pneumonia variables
orthopnea_group <- picsure::buildClauseGroup(
    lapply(orthopnea_variable_paths_of_interest, function(p) picsure::buildClause(p, type = picsure::PhenotypicFilterType$ANYRECORD)),
    operator = picsure::GroupOperator$OR
)
pneumonia_group <- picsure::buildClauseGroup(
    lapply(pneumonia_variable_paths_of_interest, function(p) picsure::buildClause(p, type = picsure::PhenotypicFilterType$ANYRECORD)),
    operator = picsure::GroupOperator$OR
)
full_group <- picsure::buildClauseGroup(list(orthopnea_group, pneumonia_group), operator = picsure::GroupOperator$AND)

query <- picsure::buildQuery(phenotypicFilter = full_group)
results <- picsure::runQuery(session, query, type = picsure::QueryType$PARTICIPANT)
```

```{r}
# Label each concept-path column by its study (derived from the phs accession in the path)
study_from_path <- function(p) {
    if (grepl('phs000007', p)) {
        'fhs'
    } else if (grepl('phs000287', p)) {
        'chs'
    } else if (grepl('phs000209', p)) {
        'mesa'
    } else {
        'unknown'
    }
}
rename_map <- c()
for (p in orthopnea_variable_paths_of_interest) rename_map[p] <- paste0(study_from_path(p), '_orthopnea')
for (p in pneumonia_variable_paths_of_interest) rename_map[p] <- paste0(study_from_path(p), '_pneumonia')

new_names <- colnames(results)
for (i in seq_along(new_names)) {
    if (new_names[i] %in% names(rename_map)) new_names[i] <- rename_map[[new_names[i]]]
}
colnames(results) <- new_names
head(results)
```

### Clean and wrangle data

```{r}
# Record study acronym
results$study <- ''
results$study[!is.na(results$chs_pneumonia)  & results$chs_pneumonia  != ''] <- 'CHS'
results$study[!is.na(results$mesa_pneumonia) & results$mesa_pneumonia != ''] <- 'MESA'
results$study[!is.na(results$fhs_pneumonia)  & results$fhs_pneumonia  != ''] <- 'FHS'
head(results)
```

```{r}
# combine orthopnea variables
results$orthopnea <- ifelse(is.na(results$fhs_orthopnea) | results$fhs_orthopnea == '', results$chs_orthopnea, results$fhs_orthopnea)
results$orthopnea <- ifelse(is.na(results$orthopnea) | results$orthopnea == '', results$mesa_orthopnea, results$orthopnea)
head(results)
```
```{r}
# combine pneumonia variables
results$pneumonia <- ifelse(is.na(results$fhs_pneumonia) | results$fhs_pneumonia == '', results$chs_pneumonia, results$fhs_pneumonia)
results$pneumonia <- ifelse(is.na(results$pneumonia) | results$pneumonia == '', results$mesa_pneumonia, results$pneumonia)
head(results)
```

As you may have noticed, the orthopnea and pneumonia variables are encoded across these studies. Let's take a look at the range of values (printed below).

```{r}
print(table(results$orthopnea))
print(table(results$pneumonia))
```

### Visualize
```{r}
results$orthopnea[is.na(results$orthopnea) | results$orthopnea == ''] <- 'DO NOT KNOW'
results$pneumonia[is.na(results$pneumonia) | results$pneumonia == ''] <- 'DO NOT KNOW'
head(results)
```

```{r}
# FHS figure
results %>% filter(study == 'FHS') %>% select(pneumonia, orthopnea) %>%
    mutate(pneumonia = ifelse(pneumonia == 'YES', 1, 0), orthopnea = ifelse(orthopnea == 'YES', 1, 0)) %>%
    as.matrix %>% vennCounts %>% vennDiagram(main = 'FHS Dataset')

# MESA figure
results %>% filter(study == 'MESA') %>% select(pneumonia, orthopnea) %>%
    mutate(pneumonia = ifelse(pneumonia == 'YES', 1, 0), orthopnea = ifelse(orthopnea == 'YES', 1, 0)) %>%
    as.matrix %>% vennCounts %>% vennDiagram(main = 'MESA Dataset')

# CHS figure
results %>% filter(study == 'CHS') %>% select(pneumonia, orthopnea) %>%
    mutate(pneumonia = ifelse(pneumonia == 'YES', 1, 0), orthopnea = ifelse(orthopnea == 'YES', 1, 0)) %>%
    as.matrix %>% vennCounts %>% vennDiagram(main = 'CHS Dataset')

# Combined figure
results %>% select(pneumonia, orthopnea) %>%
    mutate(pneumonia = ifelse(pneumonia == 'YES', 1, 0), orthopnea = ifelse(orthopnea == 'YES', 1, 0)) %>%
    as.matrix %>% vennCounts %>% vennDiagram(main = 'Combined results across FHS, MESA, and CHS')
```
