---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Connecting to BioData Catalyst HPDS resources: manualing setting up a security token


**This Notebook needs to be executed using a python kernel**


To connect to a PIC-SURE network, the PIC-SURE API needs two data: 
- The URL of this network: those notebooks examples are currently using COPDGene-dev network environment, which URL is: https://picsure.biodatacatalyst.nhlbi.nih.gov/picsure
- An individual user token, mandatory to gain access to the controlled-access data through the API.


We will hence need to get a token and provide it to the API. First, we create a blank text file called `tokens.txt` that will be used to store the token.

```{r}
file = "token.txt"
if (!file.exists(file)) file.create(file)
```

To actually get the token, process as below:


1. In a web browser, open the BioData Catalyst login page: https://picsure.biodatacatalyst.nhlbi.nih.gov/, and click on the 'eRA Commons' button to log in.
2. On the user-interface click on USER PROFILE
3. On the pop-up window, click on REFRESH and then COPY
4. Back into your Jupyter environment, paste it into the newly created text file (`./tokens.txt`).


**The user token is strictly personal, be careful not to share it with anyone**


<!-- see second answer on https://stackoverflow.com/questions/15557392/how-do-i-display-images-from-google-drive-on-a-website -->
<img src="https://drive.google.com/uc?id=1sMPtXpffcVOVVIsg8H-dpCm1VP-VcT2q">


```{r}

```
